---
title: "GradX – Graduation Success Predictor"
author: "Prathiksha Rumale Vishwanath"
output:
  pdf_document: default
  html_document: default
---

```{r install, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages("tidyverse", dependencies = TRUE)
install.packages("readr", dependencies = TRUE)
install.packages("dplyr", dependencies = TRUE)
install.packages("ggplot2", dependencies = TRUE)
install.packages("car", dependencies = TRUE)
install.packages("knitr", dependencies = TRUE)
install.packages("caret", dependencies = TRUE)
```

```{r load, include=FALSE}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(car)
library(knitr)
library(caret)
```

## Introduction: Understanding Student Dropout Risk

In this analysis, we aim to predict whether a student will **graduate or drop out** based on various academic, financial, and demographic factors. The dataset contains information about student enrollment, previous qualifications, financial aid status, academic performance, and socioeconomic indicators. By applying **logistic regression**, we aim to identify key predictors of student success and help educational institutions implement data-driven retention strategies.

**Objectives:**
- Identify the most significant factors affecting student dropout rates.
- Train and evaluate a predictive model using real student data.
- Provide actionable insights to improve student retention.

## Load the Data

```{r load-data}
# Load student enrollment data set (CSV file)
data <- read_csv("data.csv", col_names = TRUE, col_types = cols())
head(data)
```

## Data Cleaning

```{r data-cleaning}
# Check for missing values and display only columns with missing data
missing_values <- colSums(is.na(data))
missing_values[missing_values > 0]
```

```{r fix-delimiter}
# Reload data set with correct delimiter once observed and analyzing presence or missing values
data <- read.csv("data.csv", sep = ";", header = TRUE)
head(data)
```

## Student Enrollment by Course

```{r course-enrollment}
# Count number of students per course
course_enrollment <- data %>% 
  group_by(Course) %>% 
  summarise(Number_of_Students = n()) %>% 
  arrange(desc(Number_of_Students))

# Display summary of top 10 enrolled courses
top_courses <- head(course_enrollment, 10)
print(top_courses)

```

```{r plot-course-enrollment, echo=FALSE}
# Plot: Top 10 Courses with Highest Enrollment
ggplot(top_courses, aes(x = reorder(Course, -Number_of_Students), y = Number_of_Students)) +
  geom_bar(stat = "identity", fill = "pink") +
  theme_minimal() +
  labs(title = "Top 10 Courses with Highest Student Enrollment", x = "Course ID", y = "Number of Students") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Dropout vs Graduation Rate

```{r dropout-distribution}
# Count students based on their academic status
dropout_distribution <- data %>% 
  group_by(Target) %>% 
  summarise(Count = n())

# Display summary
dropout_distribution
```

```{r plot-dropout-distribution, echo=FALSE}
# Plot: Dropout vs Graduation Rate
ggplot(dropout_distribution, aes(x = Target, y = Count, fill = Target)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Student Status (Target) Distribution", x = "Student Status", y = "Count")
```

## Statistical Analysis: Predicting Dropout

```{r checking dataset column names}
#Analyzing all columns for feature engineering
colnames(data)
```

## Train-Test Split and Model Evaluation

```{r Test the Logistic Regression Model}
# Split data into 80% training and 20% testing set
set.seed(42)
trainIndex <- createDataPartition(data$Target, p = 0.8, list = FALSE)
trainData <- data[trainIndex, ]
testData <- data[-trainIndex, ]

# Ensure Target is a factor
trainData$Target <- factor(trainData$Target, levels = c("Dropout", "Graduate"))
testData$Target <- factor(testData$Target, levels = c("Dropout", "Graduate"))

# Train logistic regression model
model <- glm(Target ~ Age.at.enrollment + Scholarship.holder + Gender + 
             Curricular.units.1st.sem..grade. + Curricular.units.2nd.sem..grade., 
             data = trainData, family = binomial)
```

##Summary

```{r}
# Display model summary
summary(model)
```

## Variance Inflation Factor (VIF) Analysis

```{r}
# Check multicollinearity among predictor variables
vif(model)
```
##Confusion Matrix

```{r}
# Make predictions on test data
predictions <- predict(model, newdata = testData, type = "response")

# Convert predicted probabilities into class labels
predicted_classes <- ifelse(predictions >= 0.5, "Graduate", "Dropout")

# Evaluate model performance using confusion matrix
confusionMatrix(factor(predicted_classes), testData$Target)
```

## New Student Dropout or Graduation Prediction

```{r new-student-prediction}
# Define new student data for prediction
new_data <- data.frame(
  Age.at.enrollment = c(21, 25, 30),
  Scholarship.holder = c(1, 0, 1),  # 1 = Has Scholarship, 0 = No Scholarship
  Gender = c(1, 0, 1),  # 1 = Male, 0 = Female
  Curricular.units.1st.sem..grade. = c(14, 10, 12),
  Curricular.units.2nd.sem..grade. = c(15, 9, 11)
)

# Predict dropout probability
predictions <- predict(model, newdata = new_data, type = "response")

# Convert probabilities into class labels
predicted_classes <- ifelse(predictions >= 0.75, "Graduate", "Dropout")

# Display results
prediction_results <- data.frame(new_data, Predicted_Status = predicted_classes, Probability = predictions)
print(prediction_results)
```

## Conclusion and Actionable Insights
This analysis provides insights into **student enrollment trends and dropout rates**, helping institutions understand key factors influencing student retention.

### **Key Takeaways:**
- **VIF Analysis:** No severe multicollinearity issues detected, ensuring predictor stability.
- **Model Accuracy:** **~81.1%** accuracy with a sensitivity of **63.0%** and specificity of **92.7%**.
- **Key Predictors:** Scholarship holders have significantly higher graduation rates, while older students are more likely to drop out.
- **Predictions:** The model successfully classifies new students based on their academic and demographic factors.

### **Actionable Insights:**
1. **Scholarship Programs Significantly Reduce Dropout Rates**
   - Students who receive scholarships have a **much higher likelihood of graduating** than those without financial aid.
   - **Action:** Educational institutions should **increase scholarship availability** or introduce **financial assistance programs** to support at-risk students.

2. **Early Academic Performance is a Key Indicator of Retention**
   - **First-semester grades** strongly correlate with dropout likelihood.
   - **Action:** Universities should implement **early intervention strategies** such as **mentoring, tutoring, and academic support programs** for students struggling in their **first semester**.
